Red Wine Exploration
========================================================

What property makes good red wine? In this project we try to answer this question by exploring the red wine data set. 

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(lattice)
library(MASS)
library(gridExtra)
library(dplyr)
library(reshape)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
redwine<-read.csv('wineQualityReds.csv')
names(redwine)
```

#Univariate Plots Section

### Feature Names and Summary

This red wine data set contains 1,599 obersvations with 11 variables on the chemical properties of the wine.

```{r echo=FALSE, Univariate_selection}
names(redwine)
summary(redwine)
``` 

###  Quality Distribution

The wine quality grade is a discrete number. It is ranged from 3 to 8. The median value is at 6.

```{r echo=FALSE, Quality_Distribution}
ggplot(aes(x = quality), data = redwine)+
geom_histogram(colour="black", fill="#993366", width=.7,binwidth = 1)+
                scale_x_discrete()+
                xlab("wine grade") + ylab("count") +
                ggtitle("Distribution of Wine Quality")

```

### Distribution of Other Chemical Properties

```{r echo=FALSE,  Other_Properties}
p1<-
  ggplot(aes(x = fixed.acidity), data = redwine)+
        xlab("fixed acidity")+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 0.5)
p2<-
  ggplot(aes(x = volatile.acidity), data = redwine)+
        xlab("volatile acidity")+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 0.05)
p3<-
  ggplot(aes(x = citric.acid), data = redwine)+
        xlab("citric acid")+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 0.05)
p4<-
  ggplot(aes(x = residual.sugar), data = redwine)+
        xlab("residual sugar")+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 0.4)
p5<-
  ggplot(aes(x = chlorides), data = redwine)+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 0.025)
p6<-
  ggplot(aes(x = free.sulfur.dioxide), data = redwine)+
        xlab("free sulfur dioxide")+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 4)
p7<-
  ggplot(aes(x = total.sulfur.dioxide), data = redwine)+
        xlab("total sulfur dioxide")+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 20)
p8<-
  ggplot(aes(x = density), data = redwine)+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 0.001)
p9<-
  ggplot(aes(x = pH), data = redwine)+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 0.05)
p10<-
  ggplot(aes(x = sulphates), data = redwine)+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 0.1)
p11<-
  ggplot(aes(x = alcohol), data = redwine)+
        geom_bar(colour="black", fill="#993366", width=.7,binwidth = 0.25)

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,ncol=3)
```

### Univariate Analysis

Some observation on the distribution of the chemical property can be made:

- **Normal**:  Volatile acidity, Density, PH

- **Positively Skewed**: Fixed acidity, Citric acid, Free sulfur dioxide, Total sulfur dioxide, Sulphates, Alcohol

- **Long Tail**: Residual sugar, Chlorides

### Rescale Variable

Skewed and long tail data can be transformed toward more normally distribution by taking square root or log function. 
Take Sulphates as a example, we compare the original, square root and log of the feature.

```{r echo=FALSE,  Rescale_Variable}
# original 
p1<-
  ggplot(aes(x = sulphates), data = redwine)+
        geom_histogram(colour="black", fill="#993366", 
                       width=.7,binwidth = 0.1)

# square root
redwine$sqrt_sulphates=sqrt(redwine$sulphates)
p2<-
    ggplot(aes(x = sqrt_sulphates), data = redwine)+
          xlab("square root of sulphates")+
          geom_histogram(colour="black", fill="#993366", 
                         width=.7,binwidth = 0.05)

#log
redwine$log_sulphates=log(redwine$sulphates)
p3<-
   ggplot(aes(x = log_sulphates), data = redwine)+
          xlab("log of sulphates")+
          geom_histogram(colour="black", fill="#993366", 
                         width=.7,binwidth = 0.1)

grid.arrange(p1,p2,p3,ncol=1)
```

Both the square root and the log function helps transform the feature toward normal distribution. In comparison, the log scale feature is more normal distributed.

# Bivariate Plots Section

### Bivariate Plots Selection

Plot matrix was used to have a glance at the data. We are interested the correlation between the wine quality and each chemical property.

```{r echo=FALSE, Plot_Matrix}
# create a plot matrix
ggpairs(redwine,params=c(shape=I('.'),outlier.shap=I('.')))

```

The top 4 factors that is correlated with the wine quality (with a correlation coeffcient greater than 0.2)

| Property         | r-value |
|------------------|---------|
| alcohol          | 0.476   |
| volatile.acidity | -0.391  |
| sulphates        | 0.251   |
| citric.acid      | 0.226   |

### Bivariate Analysis

Alcohol content has the biggest correlation value to the wine quality. The scatter plot of alcohol and wine quality is shown below.


```{r echo=FALSE, Scatter_Plots_Original}
ggplot(aes(x=quality,y=alcohol),data=redwine)+
      geom_point()
```

The original plot looks over plotted, so we add alpha value and 0.1, 0.5 and 0.9 percentile line to show the general trends.

```{r echo=FALSE, Scatter_Plots_Improved}
ggplot(aes(x=quality,y=alcohol),data=redwine)+
    geom_point(color='#993366',alpha=1/4)+
    geom_line(stat='summary',fun.y = quantile,probs = .5,color='#FF6660')+
    geom_line(stat='summary',fun.y = quantile,probs = .9,
            linetype =2, color='#FF6660')+
    geom_line(stat='summary',fun.y = quantile,probs = .1,
            linetype =2, color='#FF6660')+
            xlab("Wine Grade") + ylab("Alcohol") +
    ggtitle("Wine Qaulity and Alchohol")
```

In this plot the trend of increasing wind quality with the increasing of alcohol content can be clearly observed. 

### Transforming Wine Quality into Categorical Data
Since the wine quality are desecrate value, we can transform it from numerical data to categorical data. So that box plot can be used to represent the data.

```{r echo=FALSE, Transform_Quality}
# turn quality into categroical 
redwine$grade_number<-cut(redwine$quality, 
                          c(2.5,3.5,4.5,5.5,6.5,7.5,8.5),
                          labels=c('3','4','5','6','7','8'))
```

Higher quality wine tend to have higher alcholol content.
```{r echo=FALSE, Alcholo_Boxplot}
ggplot(redwine, 
       aes(x=grade_number, y=alcohol,fill=grade_number)) + 
       geom_boxplot()+
       xlab("wine grade") + ylab("alcohol") +
       ggtitle("Alchohol vs Wine Grade")
```


Similar analysis was done the 3 other factors: volatile acidity, sulphates and citric acid

```{r echo=FALSE, More_Boxplot}

ggplot(redwine, 
       aes(x=grade_number, y=volatile.acidity,fill=grade_number)) +
       geom_boxplot()+
       xlab("wine quality") + 
       ylab("volatile acidity") +
       ggtitle("Volatile Acidity vs Wine Grade")

ggplot(redwine, 
       aes(x=grade_number, y=sulphates,fill=grade_number)) +
       geom_boxplot()+
       xlab("wine quality") + ylab("sulphates") +
       ggtitle("Sulphatesy vs Wine Grade")

ggplot(redwine, 
       aes(x=grade_number, y=citric.acid,fill=grade_number)) + 
       geom_boxplot()+
       xlab("wine quality") + ylab("citric acid") +
       ggtitle("Citric Acid vs Wine Grade")

```

### Distribution Analysis

In this analysis, we try to find out if the distribution of the chemical properties are different at different wine grade.

```{r echo=FALSE, Alchool_Grade}
ggplot(redwine, aes(x=alcohol, fill=grade_number)) +
      geom_histogram(binwidth=.5, alpha=.5, position="identity")

```

Note that sine the data size for each quality is not equal, the distribution of higher and lower grades are hard to see. 

A normalized plot is shown below.

```{r echo=FALSE, Alchool_Grade_Normalized}
ggplot(redwine, aes(x=alcohol, fill=grade_number)) +
      geom_density(aes(y=..density..),
                  binwidth=.5, alpha=.5, position="identity")+
      xlab("wine grade") + ylab("alcohol") +
      ggtitle("Alchohol Distribuiton of Differnt Wine Grade")
```

The plot looks a little busy. We group 2 grade together: grade 3,4 as "Low", grade 5,6 as "Medium", grade 7,8 as "High". And plot again.

```{r echo=FALSE, Regroup}
#3,4 low 5,6,medium 7,8high
redwine$grade<-cut(redwine$quality, c(2.5,4.5,6.5,8.5),
                   labels=c('Low','Medium','High'))

ggplot(redwine, aes(x=alcohol, fill=grade)) +
      geom_density(aes(y=..density..),binwidth=.5, 
                  alpha=.5, position="identity")+
      xlab("wine grade") + ylab("alcohol") +
      ggtitle("Alchohol Distribuiton of Differnt Wine Grade")
```

The new plot looks cleaner.

Similar analysis was done on the 3 other factors: volatile acidity, sulphates and citric acid.

```{r echo=FALSE, More_Distribution_Grade}
ggplot(redwine, aes(x=volatile.acidity, fill=grade)) +
      geom_density(aes(y=..density..),
                  binwidth=.5, alpha=.5, position="identity")+
      xlab("wine grade") + ylab("volatile acidity") +
      ggtitle("Volatile Acidity Distribuiton of Differnt Wine Grade")

ggplot(redwine, aes(x=citric.acid, fill=grade)) +
      geom_density(aes(y=..density..),
                  binwidth=.5, alpha=.5, position="identity")+
      xlab("wine grade") + ylab("citric acid") +
      ggtitle("Citric Acid Distribuiton of Differnt Wine Grade")
```

As stated in section 1 the sulphates data is skewed, we tried both the original and the log scale of the feature.


```{r echo=FALSE, log_sulphates}

ggplot(redwine, aes(x=sulphates, fill=grade)) +
      geom_density(aes(y=..density..),
                  binwidth=.5, alpha=.5, position="identity")+
      xlab("wine grade") + ylab("sulphates") +
      ggtitle("Sulphates Distribuiton of Differnt Wine Grade")

ggplot(redwine, aes(x=log_sulphates, fill=grade)) +
      geom_density(aes(y=..density..),
                   binwidth=.5, alpha=.5, position="identity")+
      xlab("wine grade") + ylab("log of sulphates") +
      ggtitle("Sulphates Distribuiton of Differnt Wine Grade")
```

The log scaled feature looks more spread out and therefore preferable.


### Median Value of Each Chemical Properties

The median value of chemical properties at each wine quality is  shown. The value is normalize by the  maximum value so that all the values ranges from 0 to 1.


```{r echo=FALSE, Rearrange}
  # rearrange the data on midian values
  redwine.median<-redwine %>%
  #  filter(!is.na(gender))%>%
  group_by(quality)%>%
  summarise(fixed.acidity = median(fixed.acidity),
            volatile.acidity = median(volatile.acidity),
            citric.acid = median(citric.acid),
            residual.sugar = median(residual.sugar),
            chlorides = median(chlorides),
            free.sulfur.dioxide = median(free.sulfur.dioxide),
            total.sulfur.dioxide = median(total.sulfur.dioxide),
            density = median(density),
            pH = median(pH),
            sulphates = median(sulphates),
            alcohol = median(alcohol),
            n=n()) %>%
  ungroup() %>%
  arrange(quality)

# normailize with maxium value
redwine.median$fixed.acidity = 
  redwine.median$fixed.acidity/max(redwine.median$fixed.acidity)
redwine.median$volatile.acidity = 
  redwine.median$volatile.acidity/max(redwine.median$volatile.acidity)
redwine.median$citric.acid = 
  redwine.median$citric.acid/max(redwine.median$citric.acid)
redwine.median$residual.sugar = 
  redwine.median$residual.sugar/max(redwine.median$residual.sugar)  
redwine.median$chlorides = 
  redwine.median$chlorides/max(redwine.median$chlorides)
redwine.median$pH=redwine.median$pH/max(redwine.median$pH)
redwine.median$free.sulfur.dioxide = 
  redwine.median$free.sulfur.dioxide/max(redwine.median$free.sulfur.dioxide)
redwine.median$total.sulfur.dioxide = 
redwine.median$total.sulfur.dioxide/max(redwine.median$total.sulfur.dioxide)
redwine.median$density = 
  redwine.median$density/max(redwine.median$density)
redwine.median$pH = redwine.median$pH/max(redwine.median$pH)
redwine.median$sulphates = 
  redwine.median$sulphates/max(redwine.median$sulphates)
redwine.median$alcohol = redwine.median$alcohol/max(redwine.median$alcohol)
```

```{r echo=FALSE, Long_Table}
# turn wide table to long table

#install.packages("reshape")
#re_arranged<-melt(redwine.median, id.vars = "quality", measure.vars = c("fixed.acidity", "volatile.acidity","citric.acid","residual.sugar","chlorides","free.sulfur.dioxide","total.sulfur.dioxide","density","pH","sulphates","alcohol"), variable_name = "abcValues")

# Install the package has a conflict with ggplot and may cause the software to crush. So here we load the data instead. The data is processed using the commented code abrove.

load("re_arranged.RData")
load("re_arranged2.RData")

```

```{r echo=FALSE, Median}
ggplot(re_arranged2, aes(x=quality, y=value, colour=properties)) + 
      geom_line(aes(group=properties)) +
      geom_point(size=3)  
```

The 4 features that has the strongest correlation to the wine quality.

```{r echo=FALSE, Median_Key}

ggplot(re_arranged, aes(x=quality, y=value, colour=properties)) + 
      geom_line(aes(group=properties)) +
      geom_point(size=3)  
```

### Correlation Between Features

There is interesting correlaiton between two of the main features: Volatile acidity and Citric acid.

```{r echo=FALSE, Correlation}
ggplot(redwine, aes(x=volatile.acidity,y=citric.acid)) +
      geom_point(color='#993366',alpha=1/2)+
      geom_smooth(se = FALSE,color='blue')+
      xlab("volatile acidity") + ylab("citric acid") +
      ggtitle("The Correlation Betwee Volatile acidity and Citric acid")

# calculate correlatio coefficient
cor.test(redwine$volatile.acidity, redwine$citric.acid,method='pearson') 

```



# Multivariate Plots Section

### Main Chemical Property vs Wine Quality

With different colors, we can  add another dimension into the plot. There are 4 main features.Alcohol, volatile acidity are the top two factor that affect wine quality. 

```{r echo=FALSE, Multivariate_Plots}
  ggplot(redwine, aes(y=alcohol,x=quality,
                      color = volatile.acidity) )+
        geom_point(alpha=1/4,size=3)+
        scale_colour_gradient2(low="red", 
                               high="blue",mid="#9933CC",midpoint = 0.8)

```

The figure looks over ploted, since the wine quality are discrete numbers. We can use jitter plot to alleviate this problem

```{r echo=FALSE, Jitter_Plots}
  ggplot(redwine, aes(y=alcohol,x=quality,color = volatile.acidity) )+
        geom_jitter(alpha=1/2,size=3)+
        xlab("alcohol") + ylab("wine quality") +
        scale_colour_gradient2(low="red", high="blue",
                               mid="#9933CC",midpoint = 0.8)

```


We can see higher quality wine have higher alcohol and lower volatile acidity.

### Add Another Feature 

Now we add the third feature, the log scale of sulphates, and use different facet to show wine grade. 

```{r echo=FALSE, Third_Features, fig.width=7, fig.height=7}

ggplot(redwine, aes(x=alcohol,y=volatile.acidity)) +
      geom_point(aes(color =  log_sulphates),alpha=1/2,size=3)+
      xlab("alcohol") + ylab("log of sulphates") +
      scale_colour_gradient2(low="red", high="blue",
                            mid="#9933CC",midpoint = -0.25)+
      facet_grid(grade ~ .) 
```

We can see higher quality wine have higher alcohol (x-axis), lower volatile acidity (y-axis) and higher sulphates (hue).

### Main Chemical Properties  vs Wine Quality

Since we can visualized 3 dimensions, including wine quality, at a time. Two graphs will be needed to visualize the 4 main chemical properties.

```{r echo=FALSE, Main_Chemical_Property}

ggplot(aes(x = volatile.acidity, y = alcohol), data = redwine) +
      xlab("volatile acidity") + ylab("alcohol") +
      geom_point(aes(color =  grade),size=2)
```

The same trend of alcholand volatile acidity's effect on wine qaulity can be observed.

```{r echo=FALSE, Main_Chemical_Property2}

ggplot(aes(x = log_sulphates, y = citric.acid), data = redwine) +
      xlab("log sulphates") + ylab("citric acid") +
      geom_point(aes(color =  grade),size=2)
```

We can see higher quality wine have higher sulphates (x-axis), higher citric acidity (y-axis).

# Linear Multivariable Model

Linear multivariable model was created to predict the wine quality based on chemical properties. 

The features are selected incrementally in order of how strong the correlation between this feature and wine quality.  

```{r echo=FALSE, Linear_Model}

  # regression
  m1<-lm(quality ~ volatile.acidity,data=redwine)
  m2<-update(m1,~. + alcohol)
  m3<-update(m2,~. + sulphates)
  m4<-update(m3,~. + citric.acid)
  m5<-update(m4,~. + chlorides)
  m6<-update(m5,~. + total.sulfur.dioxide)
  m7<-update(m6,~. + density)
  mtable(m1,m2,m3,m4,m5,m6,m7)

```

The model of 6 features has the lowest AIC (Akaike information criterion) number. As the number of features increase the AIC becomes higher. The parameter of the predictor also changed dramatically which shows a sign of overfitting.

The model can be described as:

wine_quality = 2.985 + 0.276xalcohol - 2.985xvolatile.acidity + 0.908xsulphates + 0.065xcitric.acid - -1.763*chlorides - 0.002xtotal.sulfur.dioxide 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
  ggplot(aes(x = quality), data = redwine)+
  geom_histogram(colour="black", fill="#993366", width=.7,binwidth = 1)+
                 scale_x_discrete()+
                xlab("wine grade") + ylab("count") +
                ggtitle("Distribution of Wine Quality")

```

### Description One

Plot one shows the distribution of wine quality. Note that dataset is unbalanced. It has many count for medium quality (grade 5, 6), but much fewer count on low (grade 3,4) and high (grade 7, 8) quality wine. 


### Plot Two: 
```{r echo=FALSE, Plot_Two}
p1<-
ggplot(redwine, aes(x=grade, y=alcohol,fill=grade)) +
      geom_boxplot()+
      xlab("wine quality") + ylab("alcohol") +
      ggtitle("Alcohol vs Wine quality")

p2<-
ggplot(redwine, aes(x=grade, y=volatile.acidity,fill=grade)) +
      geom_boxplot()+
      xlab("wine quality") + 
      ylab("volatile acidity") +
      ggtitle("volatile acidity vs wine grade")

p3<-
ggplot(redwine, aes(x=grade, y=sulphates,fill=grade)) +
      geom_boxplot()+
      xlab("wine quality") + ylab("sulphates") +
      ggtitle("Sulphatesy vs Wine grade")

p4<-
ggplot(redwine, aes(x=grade, y=citric.acid,fill=grade)) +
      geom_boxplot()+
      xlab("wine quality") + ylab("citric") +
      ggtitle("Citric acid vs Wine grade")
      grid.arrange(p1,p2,p3,p4,ncol=2)

```

### Description Two

The 4 features that have the highest correlation coefficient are alcohol, volatile acidity, sulphates, citric acid. The wine quality are grouped to low (3,4) medium (5.6) and high (7,8). High quality wine have high alcohol level however, there is no significant different between medium and low quality wine. Volatile acidity decrease as wine quality increases. Sulphate and critic acid increase as wine quality increase.


### Plot Three
```{r echo=FALSE, Plot_Three, fig.width=7, fig.height=9}

p1<-
ggplot(aes(x = volatile.acidity, y = alcohol), data = redwine) +
      geom_point(aes(color =  grade),size=2)+
      xlab("volatile acidity") + ylab("alcohol") +
      ggtitle("Volatile acidity and Alcohol")

p2<-
ggplot(aes(x = log_sulphates, y = citric.acid), data = redwine) +
      geom_point(aes(color =  grade),size=2)+
      xlab("log of sulphates") + ylab("citric acid") +
      ggtitle("Citric Acid and Sulphates")

grid.arrange(p1,p2,ncol=1)


```

### Description Three

The 4 features are also represented in the scatter plot. 2 features are plotted at a time with color indicate wine quality. Similar trend as the last figure can be observed. In general, high quality wine tend to have higher alcohol and lower volatile acidity content. They also tend to have higher sulphate and higher critic acid content.

------

# Reflection

The red wine dataset  contains 1,599 observation with 11 variables on the chemical properties. We are interested in the correlation between the features and wine quality. Unlike the diamond price,  which is the  dominated by their size or carat. The wine quality is more complex. It does not have a obvious driver. Most of the data visualization in this project was done on the 4 features that have the highest correlation coefficient: alcohol(0.476), volatile acidity(-0.391), sulphates(0.251),citric acid(0.226). After some web research, the reflection about these chemical component are as follows. 

 - **Alcohol**: surprisingly and unsurprisingly, alcohol is the No.1 factor correlated to the wine quality. The data strongly suggest that the higher the alcohol content, the more likely the better wine quality. One suggestion is that wine of higher alcohol are made from riper grapes, which tend to have intense flavor. Therefore, the relation between alcohol and wine quality are more likely to be correlation rather than causation. There is also controversy about alcohol level. One article even says “high alcohol is a wine fault not a badge of honor”. [1][2]

 - **Volatile acidity**:  volatile acidity has a negative correlation to wine quality. Volatile acidity can contributed to acidic tastes which is often considered a wine fault.[3]

 - **Sulphates**: sulphates has a positive correlation with wine quality. It is often added by  winemakers to prevent spoilage. It is less likely that sulphates itself contribute to better taste or aroma. Its present simply means the wine is less likely to be spoiled.[4]

  - **Citic acids**: unlike volatile acid, citic acids has positive correlation with wine quality . Winemaker often add citric acid to give a “freshness” test. However it can also bring unwanted effects through bacteria metabolism.[5]

Surprisingly, other chemical proprieties do not have strong correlation with wine quality, such as the residual sugar and PH .

In the end, a linear model of 6 features was created to predict wine quality. However, wine quality is a complex object.  Different type grape can largely affect the wine test. There are many nuance in taste and aroma the that cannot be capture by examine its chemical component. The linear model is a overly simplified model. Good wine is more than perfect combination of different chemical components.

Future improvement can be made if more data can be collected on both low-quality and high-quality wine. I noticed that the dataset is highly unbalanced. It has many data points for medium quality wine (grade 5, 6). However, for low quality (grade 3,4 ) and high quality (grade 7, 8) wine, it has fewer data points. If the data set has more records on both the low end and high end, the quality of analysis can be improved. We can be more certain about whether there is a significant correlation between a chemical component and the wine quality.

### Reference

[1] http://www.realsimple.com/holidays-entertaining/entertaining/food-drink/alcohol-content-wine

[2] http://www.winegeeks.com/articles/85/high_alcohol_is_a_wine_fault_not_a_badge_of_honor/

[3] http://en.wikipedia.org/wiki/Wine_fault

[4] http://en.wikipedia.org/wiki/Sulfite

[5] http://www.calwineries.com/learn/wine-chemistry/wine-acids/citric-acid

